### 2025-09-18 paper/sec/introduction.tex
Rehash para 1.

Making sense of and verifying data-driven claims is hard, even when the evidence base lives in the public
domain, because of the difficulty in tracing specific claims back to the data. Peer review of empirical claims
without author-supplied traceability to the underlying data can be challenging~\citep{weber20}; paper
retractions are often attributable to (non-fraudulent but significant) errors in data management and
analysis~\citep{hu25}. The use of LLMs to assist with these challenges has received considerable attention
recently, on problems ranging from fact-checking~\citep{abu-ahmad25} to interpretation of charts and
figures~\citep{roberts24}, LLM interfaces do not support direct interrogation of visual outputs.

Recent work on data provenance and data visualisation~\citep{psallidas18smoke,perera22,bond25} approaches the
transparency problem from this direction, through software infrastructure which transparently connections
charts and figures to their data sources. \rpnote{Contrast with LLM-based approach -- trusted infrastructure.}
This enables visual outputs to provide interactive \emph{provenance queries}, user interactions (e.g. mousing
over a visual element) that allow readers to explore how visual features relate to data. The value proposition
of these approaches is that the interactions are provided automatically via trusted infrastructure, namely an
underlying programming language (say query language or general-purpose language) which tracks dependencies as
data flows through a computation. However, these approaches are limited to outputs computed from data, which
usually precludes natural language, despite the fact that the main claims of a paper are usually made using
natural language, with charts and figures only playing a supporting role.

>>>>

Making sense of and verifying data-driven claims is hard, even when the evidence base lives in the public
domain. A key challenge lies in tracing specific claims back to the underlying data. In peer review, for
example, empirical claims typically lack author-supplied links to data, making it hard for reviewers to check
them directly~\citep{weber20}. Paper retractions, meanwhile, are often attributable not to fraud, but to
simple errors in data management or analysis~\citep{hu25}. Large language models (LLMs) can assist with
fact-checking~\citep{abu-ahmad25} and interpretation of charts and figures~\citep{roberts24}, but currently
LLM interfaces do not support direct interrogation of visual outputs.

Recent work on data provenance and data visualisation~\citep{psallidas18smoke,perera22,bond25}, on the other
hand, approaches the transparency problem from this direction, through software infrastructure which
transparently connections charts and figures to their data sources. \rpnote{Contrast with LLM-based approach
-- trusted infrastructure.} This enables visual outputs to provide interactive \emph{provenance queries}, user
interactions (e.g. mousing over a visual element) that allow readers to explore how visual features relate to
data. The value proposition of these approaches is that the interactions are provided automatically via
trusted infrastructure, namely an underlying programming language (say query language or general-purpose
language) which tracks dependencies as data flows through a computation. However, these approaches are limited
to outputs computed from data, which usually precludes natural language, despite the fact that the main claims
of a paper are usually made using natural language, with charts and figures only playing a supporting role.
